 ---
title: "QRM2"
author: "Groepje 47"
date: "2026-09-04"
output: html_document
---
# Week 1

## 1 Dataset

Find the dataset “movies5.tsv” on Canvas. Describe your data set: How many observations does it have? How many variables are there? How many subjects? What consists of a subject? [4 points]

```{r}
# Load the dataset
movies <- read.delim("movies5.tsv")

# Count the rows (observations)
nrow(movies)
# Count the columns (variables)
ncol(movies)

# Count the unique movie identifiers
length(unique(movies$index))
```

**Your Answer:**

Number of observations: 909 observations (rows).

Number of variables: 19 variables (columns).

Number of subjects: 909 subjects (films)

The dataset contains 909 observations and 19 variables, including the identifier “index”. There are 909 subjects, each representing an individual movie. Each row contains one observation with information about a movie, such as its title, budget, revenue, and runtime etc. Since each movie appears exactly once in the dataset, the number of observations equals the number of subjects.

## 2 Types of variables

Which of the following types of variables are present in your data set? (i) nominal; (ii) ordinal; (iii) interval; (iv) ratio. If present, name one example of such a variable present in your data set. [4 points]

```{r}
# Show the variables and their data types
str(movies)

# Show the different genres
unique(movies$genre)

# Show the different ratings from lowest to highest
sort(unique(movies$vote_average))

# Show the first six release years
head(movies$release_year)
# Show the first six runtimes
head(movies$runtime)
```

**Your Answer:**

(i) “Genre” is a nominal variable, since categories like Comedy, Drama, and Action don’t have any natural order.

(ii) The dataset does not include an ordinal variable. We treat “vote_average” as an interval variable, which assumes the rating points have a even distance apart.

(iii) “Release_year” qualifies as an interval variable. The gap between two years always means the same thing—ten years is ten years—yet there’s no true zero. So, moving from 2000 to 2010 is just as much time as 2010 to 2020.

(iv) “Runtime” is a ratio variable. Time has a true zero, and we can talk about its values in terms of multiples—a 120-minute movie is twice as long as a 60-minute one.

## 3 Movie profits and ratings

A movie studio wants to know which types of movies give maximal profit. Perform the following steps to provide the movie studio with an analysis which corresponds to their request:

a. Create the variable profits as the revenue of a movie minus its budget. Report its mean, median, maximum, and minimum. [2 points]

b. Which movie has the highest profits in your data set and how much are these profits? Which movie has the lowest and how much are its profits? If multiple movies have the exact same highest or lowest profits, give only one example. [2 points]

c. Create a boxplot of the variable profits. Make sure it has an appropriate title, and appropriate titles and labels for the x- and y-axis. Give Q1, Q2, Q3, and Q4. What does this tell you about the nature of making money in the movies industry? [2 points]

d. Add a new variable to your data set the log of profits. When creating this variable, what happens to movies for which profits is zero or negative? What then happens when you calculate the mean of log of profits? [2 points]

e. For movies that have a profit of zero or less, replace log of profits with “NA”. What is now the mean of log of profits? Create a boxplot for log of profits, again with an appropriate title, x- and y-axis labels. How does it compare to the boxplot you made under c.)? [2 points]

f. Create a scatterplot of with the runtime of movies on the x-axis and the average vote of movies on the y-axis. What do you conclude from the scatterplot? Are movies with a longer runtime considered worse or better by the audience, or does the audience not have a preference? Why do you think this is the case? [2 points]

For each step, you should provide first all the code you used to answer the question and then formulate an answer using full sentences.

### Step a

```{r}
# Create profits variable
movies$profits <- movies$revenue - movies$budget

# Calculate mean/median/max/min
# Calculate the average profit
mean(movies$profits)
# Calculate the median profit
median(movies$profits)
# Show the highest profit
max(movies$profits)
# Show the lowest profit
min(movies$profits)
```

**Your Answer:**

We calculated profit by subtracting the budget from the revenue. The mean profit is 56,284,649.70, and the median is 2,942,221. The highest profit is 1,082,730,962, and the lowest is −98,301,101.

### Step b

```{r}
# Movie with the highest profit and its profit
movies$title[which.max(movies$profits)]
# Show the highest profit
max(movies$profits)

# Movie with the lowest profit and its profit
movies$title[which.min(movies$profits)]
# Show the lowest profit
min(movies$profits)
```

**Your Answer:**

“Minions” has the highest profit, earning 1,082,730,962 above its budget. “The 13th Warrior” has the lowest profit, with a loss of 98,301,101. This result was also found with max/min in step a.

### Step c

```{r}
# Boxplot of profits in millions
boxplot(movies$profits / 1000000,
        main = "Movie profits",
        xlab = "All movies",
        ylab = "Profit (millions)",
        col = "lightblue")

# Summary of profits in the original monetary units
summary(movies$profits)
```

**Your Answer:**

Q1 is −1,260,859, Q2 is 2,942,221, and Q3 is 61,340,801. We interpret Q4 as the maximum, which is 1,082,730,962.

The boxplot shows that a few movies make much higher profits than the rest, which pulls the mean above the median. Based on the recorded revenue and budget, at least a quarter of the movies made a loss. This suggests that making movies can be risky: some earn huge profits, while many others earn much less or lose money.

### Step d

```{r}
# Create the natural logarithm of profits
movies$log_profits <- log(movies$profits)

# Count movies with zero or negative profits
# Count movies with zero profit
sum(movies$profits == 0)
# Count movies with negative profit
sum(movies$profits < 0)

# Calculate the mean of log profits
mean(movies$log_profits)
```

**Your Answer:**

For movies with zero profit, taking the logarithm gives −Inf. For movies with negative profit, it gives NaN, and R shows a warning. There are 162 movies with zero profit and 261 with negative profit. Because the log-profit variable contains these values, it means the calculation of the mean becomes NaN. You would have to filter out these NaN and -Inf to correctly calculate the mean of movies with more than 0 profit.

### Step e

```{r}
# Replace log profits with NA for non-positive profits
movies$log_profits[movies$profits <= 0] <- NA

# Calculate the mean, excluding missing values
mean(movies$log_profits, na.rm = TRUE)

# Create a boxplot of log profits
boxplot(
  movies$log_profits,
  main = "Distribution of Log Movie Profits",
  xlab = "Movies",
  ylab = "Natural logarithm of profit",
  names = "Positive-profit movies",
  col = "lightgreen"
)
```

**Your Answer:**

After replacing the log values for zero and negative profits with NA, the mean log profit is about 17.5448. We used na.rm = TRUE to leave out the missing values, so this mean only includes the 486 movies with positive profits.

The new boxplot is less stretched towards high values because the logarithm brings large profits closer together. There are still several outliers on the lower side. The plots also include different movies: the first includes all movies, while the second only includes movies that made a positive profit.

### Step f

```{r}
# Create a scatterplot of runtime and average rating
plot(
  movies$runtime,
  movies$vote_average,
  main = "Movie Runtime and Average Audience Rating",
  xlab = "Runtime (minutes)",
  ylab = "Average audience rating",
  pch = 19,
  col = rgb(0, 0.3, 0.7, 0.4)
)

# Calculate the correlation to support the interpretation
cor(movies$runtime, movies$vote_average)
```

**Your Answer:**

The scatterplot suggests that longer movies tend to receive slightly higher ratings. The correlation is about 0.32, but the points are spread out, so there are plenty of exceptions.

Longer movies might have more time to develop their characters and stories, which could help explain the higher ratings. Genre and movie quality could also play a role. The plot does not prove that making a movie longer will improve its rating. There are also some zero values for runtime and rating that would need checking before drawing firm conclusions.

# Week 2

## 1 Population and sample

Is your dataset movies5.tsv the full population, or is it a sample of a larger population? If the latter, how would you describe the full population? [4 points]

**Your Answer:**

The dataset is a sample of a larger population because it contains 909 movies, rather than every movie. The full population would be all movies we want to draw conclusions about. For example, if we want to study movies released between 1990 and 2016, the population would include all movies released during those years. We cannot assume that the movies in our dataset were selected randomly.

## 2 Actors and movie revenue

### Step a

For which actor in your data set do you observe the most movies? [2 points]

```{r}
# Count the movies for each actor
actor_counts <- table(movies$first_actor)

# Show the highest counts first
head(sort(actor_counts, decreasing = TRUE))
```

**Your Answer:**

Bruce Willis and Tom Hanks appear most often, with 8 movies each. We use Bruce Willis for the next steps. These counts are based on the first_actor column, so they do not include appearances by actors elsewhere in a movie's cast.

### Step b

What is the average revenue of the movie in which this actor plays and does the revenue lie above or below the revenue of an average movie according to your data set? [2 points]

```{r}
# Average revenue of movies with Bruce Willis as first_actor
mean(movies$revenue[movies$first_actor == "Bruce Willis"],
     na.rm = TRUE)

# Average revenue of all movies
mean(movies$revenue, na.rm = TRUE)
```

**Your Answer:**

The average revenue of the movies with Bruce Willis is 228,062,987.38. The average revenue across all movies in the dataset is 88,846,314.90. His movies therefore earn more revenue on average than the average movie in this dataset.

### Step c

How trustworthy do you consider your conclusion to answer 2b? Use the term "law of large numbers" in your explanation. [2 points]

```{r}
# Number of movies used for the Bruce Willis average
sum(movies$first_actor == "Bruce Willis", na.rm = TRUE)
```

**Your Answer:**

The comparison is correct for this dataset, but it is based on only 8 Bruce Willis movies. One or two very successful movies could have a large effect on his average.

The law of large numbers tells us that, with a sufficiently large random sample, the sample mean tends to get closer to the population mean. Since we only have 8 movies and do not know whether they were randomly selected, we should be careful about applying this result to all Bruce Willis movies. It also does not show that Bruce Willis himself causes higher revenue.

## 3 Population and sample variance

For this question, you will assume that your data set is the full population.

### Step a

Recode profits such that it is expressed in millions. What is the variance of the variable profits (in millions) in your data set? [2 points]

```{r}
# Calculate profits in millions
movies$profits <- (movies$revenue - movies$budget) / 1000000

# Calculate the population variance
population_variance <- mean(
  (movies$profits - mean(movies$profits))^2
)

# Show the population variance
population_variance
```

**Your Answer:**

The population variance of profits is approximately 17,802.62. Since profits are expressed in millions, we expressed the variance in squared millions. The reason we did this is so the negative and positive would otherwise cancel out and larger differences get more weight.

We calculated the average squared distance from the mean. This divides by the number of movies, because we are treating the dataset as the full population, as said in the assignment. The usual var() function divides by the number of observations minus one and is used to calculate sample variance.

### Step b

Create a new data set, called movies_sample. Make sure that it is a random sample of your data set of 25 movies. What is the variance of profits in this random sample? How does it compare to the variance of profits in 2a? [2 points]

```{r}
# Make the random sample reproducible
set.seed(47)

# Select a random sample of 25 movies
movies_sample <- movies[sample(nrow(movies), 25), ]

# Calculate the sample variance
var(movies_sample$profits)
```

**Your Answer:**

The sample variance is approximately 15,728.93 squared millions. This is lower than the population variance of 17,802.62 from the previous step.

The values differ because the sample contains only 25 of the 909 movies. A different sample could give a higher or lower variance, depending on which movies are selected. A sample means a different variances, based on which movies you pick.

### Step c

In a for loop, create 100 different samples of 25 movies, as in b, and estimate the variance within each sample. Save the variance of each sample in a vector called sample_vars. So the first position of the vector would have the variance of the first sample, the second position the variance of the second sample, etc. Print the start of this vector. [2 points]

```{r}
# Make the random samples reproducible
set.seed(47)

# Create space for 100 sample variances
sample_vars <- numeric(100)

# Take 100 samples and save their variances
for (i in 1:100) {
  # Select 25 movies at random for this sample
  movies_sample <- movies[sample(nrow(movies), 25), ]
  # Save this sample variance in position i
  sample_vars[i] <- var(movies_sample$profits)
}

# Show the first six sample variances
head(sample_vars)
```

**Your Answer:**

We took 100 random samples of 25 movies and saved each sample variance in sample_vars. The first six variances are approximately 15,728.93, 31,971.87, 16,532.82, 28,257.89, 24,848.85, and 56,289.00.

The variances differ because each sample contains a different selection of movies, as explained in step 3b.

### Step d

Summarize and make a histogram of sample_vars. What is the mean, standard deviation and shape of its distribution? [2 points]

```{r}
# Summarize the sample variances
summary(sample_vars)

# Calculate the mean and standard deviation
# Calculate the average of the sample variances
mean(sample_vars)
# Calculate the standard deviation of the sample variances
sd(sample_vars)

# Create a histogram
hist(sample_vars,
     main = "Sample variances of movie profits",
     xlab = "Sample variance (squared millions)",
     ylab = "Frequency",
     col = "lightblue")
```

**Your Answer:**

The mean of the 100 sample variances is approximately 19,982.63, and the standard deviation is approximately 16,853.97. Both are expressed in squared millions.

The histogram shows many lower sample variances and a smaller number of much higher values, creating a long tail to the right. Samples that include movies with exceptionally high profits can have much larger variances than samples without those movies, because their profits are far above the sample mean. These large differences are squared, giving them a strong effect on the result.

### Step e

In your opinion, is a sample of 25 movies sufficient to get a reliable estimate of the population variance of profits, using the sample variance? Explain. [2 points]

**Your Answer:**

A sample of 25 movies does not seem large enough to give a reliable estimate of the population variance in this dataset. The sample variances differ a lot, and their standard deviation is large compared with their mean.

A few movies have exceptionally high profits. Including or missing one of these movies can strongly change the sample variance. A larger random sample would generally give a more stable estimate of the population variance, in this data-set.
 


