---
title: "QRM2"
author: "Groepje 47"
date: "2026-09-04"
output: html_document
---
1 Week 1
1. Find the dataset “movies5.tsv” on Canvas. Describe your data set: How many observations does it
have. How many variables are there? How many subjects? What consists of a subject? [4 points]
#WRITE YOUR CODE HERE
```{r}
movies <- read.delim("movies5.tsv")

nrow(movies)
ncol(movies)

length(unique(movies$index))
```
# Your Answer:
Number of observations: 909 observations (rows).

Number of variables: 19 variables (columns).

Number of subjects: 909 subjects (films)

The dataset contains 909 observations and 19 variables, including the identifier “index”. There are 909 subjects, each representing an individual movie. Each row contains one observation with information about a movie, such as its title, budget, revenue, and runtime etc. Since each movie appears exactly once in the dataset, the number of observations equals the number of subjects.

2. Which of the following types of variables are present in your data set? (i) nominal; (ii) ordinal; (iii);
interval; (iv) ratio. If present, name one example of such a variable present in your data set. [4 points]
#WRITE YOUR CODE HERE
```{r}


str(movies)

unique(movies$genre)

sort(unique(movies$vote_average))


head(movies$release_year)
head(movies$runtime)

```
Your Answer:
(i) A nominal variable is “genre”, because categories such as Comedy, Drama, and Action have no natural order.

(ii) There are no clearly ordinal variables in the dataset. We treat “vote_average” as an interval variable, assuming equal distances between rating points.

(iii) An interval variable is “release_year”, because equal differences represent equal amounts of time, but there is no absolute zero. For example, the difference between 2000 and 2010 is the same as between 2010 and 2020.

(iv) A ratio variable is “runtime”, because duration has an absolute zero and ratios are meaningful. For example, a movie lasting 120 minutes is twice as long as a movie lasting 60 minutes. 




3. A movie studio wants to know which types of movies give maximal profit. Perform the following steps
to provide the movie studio with an analysis which corresponds to their request:
a. Create the variable profits as the revenue of a movie minus its budget. Report its mean, median,
maximum, and minimum. [2 points]
b. Which movie has the highest profits in your data set and how much are these profits. Which movie
has the lowest and how much are its profits? If multiple movies have the exact same highest or lowest
profits, give only one example. [2 points]
c. Create a boxplot of the variable profits. Make sure it has an appropriate title, and appropriate titles
and labels for the x- and y-axis. Give Q1, Q2, Q3, and Q4. What does this tell you about the nature of
making money in the movies industry? [2 points]
d. Add a new variable to your data set the log of profits. When creating this variable, what happens to
movies for which profits is zero or negative? What then happens when you calculate the mean of log of
profits? [2 points]
e. For movies that have a profit of zero or less, replace log of profits with “NA”. What is now the mean of
log of profits? Create a boxplot for log of profits, again with an appropriate title, x- and y-axis labels.
How does it compare to the boxplot you made under c.)? [2 points]
f. Create a scatterplot of with the runtime of movies on the x-axis and the average vote of movies on the
y-axis. What do you conclude from the scatterplot? Are movies with a longer runtime considered worse
or better by the audience, or does the audience not have a preference? Why do you think this is the
case? [2 points]
For each step, you should provide first all the code you used to answer the question and then formulate an
answer using full sentences.


Step a

```{r}
# Read the dataset
movies <- read.delim("movies5.tsv")

# Create the profits variable
movies$profits <- movies$revenue - movies$budget

# Calculate the descriptive statistics
mean(movies$profits)
median(movies$profits)
max(movies$profits)
min(movies$profits)
```

Your Answer:

We calculated profit by subtracting the budget from the revenue. The mean profit is 56,284,649.70, and the median is 2,942,221. The highest profit is 1,082,730,962, and the lowest is −98,301,101. All amounts use the monetary units in the dataset.

Step b

```{r}
# Find the movie with the highest profit
movies[which.max(movies$profits), c("title", "profits")]

# Find the movie with the lowest profit
movies[which.min(movies$profits), c("title", "profits")]
```

Your Answer:

“Minions” has the highest profit, earning 1,082,730,962 above its budget. “The 13th Warrior” has the lowest profit, with a loss of 98,301,101.

Step c

```{r}
# Create a boxplot of profits
boxplot(
  movies$profits / 1000000,
  main = "Distribution of Movie Profits",
  xlab = "Movies",
  ylab = "Profit (millions of monetary units)",
  names = "All movies",
  col = "lightblue"
)

# Add a reference line at zero profit
abline(h = 0, col = "red", lty = 2)

# Calculate quartiles, interpreting Q4 as the maximum
quantile(
  movies$profits,
  probs = c(0.25, 0.50, 0.75, 1.00)
)
```

Your Answer:

Q1 is −1,260,859, Q2 is 2,942,221, and Q3 is 61,340,801. We interpret Q4 as the maximum, which is 1,082,730,962.

The boxplot shows that a few movies make very large profits. These movies pull the mean well above the median, making the distribution right-skewed. At least a quarter of the movies lose money based on their recorded revenue and budget. Making movies can therefore be risky: some earn huge profits, while many others earn much less or make a loss.

Step d

```{r}
# Create the natural logarithm of profits
movies$log_profits <- log(movies$profits)

# Count movies with zero or negative profits
sum(movies$profits == 0)
sum(movies$profits < 0)

# Calculate the mean of log profits
mean(movies$log_profits)
```

Your Answer:

For movies with zero profit, taking the logarithm gives −Inf. For movies with negative profit, it gives NaN, and R shows a warning. There are 162 movies with zero profit and 261 with negative profit. Because the log-profit variable contains these values, its mean is NaN.

Step e

```{r}
# Replace log profits with NA for non-positive profits
movies$log_profits[movies$profits <= 0] <- NA

# Calculate the mean, excluding missing values
mean(movies$log_profits, na.rm = TRUE)

# Create a boxplot of log profits
boxplot(
  movies$log_profits,
  main = "Distribution of Log Movie Profits",
  xlab = "Movies",
  ylab = "Natural logarithm of profit",
  names = "Positive-profit movies",
  col = "lightgreen"
)
```

Your Answer:

After replacing the log values for zero and negative profits with NA, the mean log profit is about 17.5448. We used `na.rm = TRUE` to leave out the missing values, so this mean only includes the 486 movies with positive profits.

The new boxplot is less stretched towards high values because the logarithm brings large profits closer together. There are still several outliers on the lower side. The plots also include different movies: the first includes all movies, while the second only includes movies that made a positive profit.

Step f

```{r}
# Create a scatterplot of runtime and average rating
plot(
  movies$runtime,
  movies$vote_average,
  main = "Movie Runtime and Average Audience Rating",
  xlab = "Runtime (minutes)",
  ylab = "Average audience rating",
  pch = 19,
  col = rgb(0, 0.3, 0.7, 0.4)
)

# Calculate the correlation to support the interpretation
cor(movies$runtime, movies$vote_average)
```

Your Answer:

The scatterplot suggests that longer movies tend to receive slightly higher ratings. The correlation is about 0.32, but the points are spread out, so there are plenty of exceptions.

Longer movies might have more time to develop their characters and stories, which could help explain the higher ratings. Genre and movie quality could also play a role. The plot does not prove that making a movie longer will improve its rating. There are also some zero values for runtime and rating that would need checking before drawing firm conclusions.
