---
title: "QRM2"
author: "Alicia Djidda (2865518), Gijs Stevens (2811736), Seeger Tromp (2889983)"
date: "2026-10-11"
documentclass: article
papersize: a4
fontsize: 10pt
geometry: margin=1.5cm
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
header-includes:
  - \setlength{\parskip}{3pt}
  - \AtBeginEnvironment{Shaded}{\small}
---

```{r setup, include=FALSE}
# Use the same clear layout for all graphs and code output
knitr::opts_chunk$set(fig.width = 6.2, fig.height = 3.0,
                      dpi = 200, collapse = TRUE, error = FALSE)
options(width = 75)
```
```{r}
# Load the dataset
library(tidyverse)
movies <- read.delim("movies5.tsv")
```



# Week 4
1. There is another argument going on in the movie studio. Bob claims that production budgets are
getting out of hand, and that the studio should focus on making cheaper movies. Chantal disagrees.
She tells Bob that “Every dollar we spend on movie production is more than offset by the increase in
movie profits’’.
6
a. Set up a regression model to test Chantal’s claim, and estimate it. That is, estimate:
Profitsi = β0 + β1Budgeti + εi.
Print a summary of your estimated model. [2 points]
b. What is the estimated value of β1 and how do you interpet it? [2 points]
c. Test for the null hypothesis that β1 ≥ 0. Report the p-value and state your conclusion. [2 points]
d. Next, estimate the model
Log Profitsi = β0 + β1Log Budgeti + εi.
When creating the variables Log Profits and Log Budget, make sure that movies with a Revenue or
Budget of zero are assigned the value “NA”. Print a summary of your estimated model [2 points]
e. What is the estimated value of β1 and how do you interpet it? [2 points]
f. Which model has better fit? The level-level model or the log-log model? Explain. [2 points]
g. Who do you think is correct? Bob or Chantal? What would you advise the movie studio to do? [2
points]
step a
# Week 4 - Question 1a: Bob vs. Chantal

```{r}
# Calculate movie profits
movies$profit <- movies$revenue - movies$budget

# Estimate the regression
m1 <- lm(profit ~ budget, data = movies)
summary(m1)
```
Write your formulated response here.
step b
#WRITE YOUR CODE HERE
Your Answer:
Write your formulated response here.
step c
```{r}
t_val <- coef(summary(m1))["budget","t value"]
pt(t_val, df = m1$df.residual, lower.tail = TRUE)   # one-sided p-value for H0: beta1 >= 0
```
Your Answer:
Write your formulated response here.
step d
```{r}
# Create log variables
movies$log_profit <- NA_real_
movies$log_budget <- NA_real_

# Select movies with positive profit, revenue and budget
valid <- with(movies, which(
  profit > 0 & revenue > 0 & budget > 0
))

movies$log_profit[valid] <- log(movies$profit[valid])

# Calculate log budget for positive budgets
valid_budget <- which(movies$budget > 0)
movies$log_budget[valid_budget] <-
  log(movies$budget[valid_budget])

# Estimate the log-log model
m2 <- lm(log_profit ~ log_budget, data = movies)
summary(m2)
```
Your Answer:
We estimated a regression of log profit on log budget. The estimated intercept is 7.64769 and the coefficient of log budget is 0.58680. The model uses 463 movies with positive profits and budgets. Its R-squared is 0.271, meaning that the model explains approximately 27.1% of the variation in log profit among these movies.
step e
```{r}
coef(m2)["log_budget"]
```
Your Answer:
The estimated coefficient of log budget is 0.58680. Since this is a log-log model, a 1% higher budget is associated with approximately 0.59% higher profit, on average. This relationship applies to the movies included in the model and does not prove that increasing a movie’s budget causes its profit to increase.

step f
```{r}
# R-squared of the level-level model
summary(m1)$r.squared

# R-squared of the log-log model
summary(m2)$r.squared

# Number of observations used in each model
nobs(m1)
nobs(m2)
```
Your Answer:
The first model explains about 32.3% of the variation in profit. The second model explains about 27.1% of the variation in log profit.

The first model has a higher R-squared, but this does not necessarily mean it fits better. One model uses profit, while the other uses log profit. The first model also uses 909 movies, compared with 463 in the second model. Because of these differences, we cannot decide which model fits better just by comparing their R-squared values.

7
step g
#WRITE YOUR CODE HERE
Your Answer:
Write your formulated response here.
2
a. Make a plot with a 95% confidence interval with the mean log of budget on the y-axis, and whether the
first actor of the movie is male or female on the x-axis. What do you conclude? [2 points]
b. Estimate the following simple OLS model: log(budget)i = β0 + β1(FirstActorMale)i + εi. Is the
estimated coefficient for β1 significantly different from zero? How do you interpret its estimate, and
how does this relate to your conclusion in 2a? [2 points]
c. Now, have a close look at your data frame. Can you find any instances of male first actors who are
wrongly labeled as being female, or vice versa? What would such mislabelling mean for the coefficient
you estimated under 2b? [2 points]
step a
```{r}
movies <- movies %>%
  mutate(log_budget = ifelse(budget > 0, log(budget), NA))

movies %>% filter(!is.na(log_budget), !is.na(first_actor_gender)) %>%
  ggplot(aes(first_actor_gender, log_budget)) +
  stat_summary(fun.data = mean_cl_normal, geom = "pointrange") +
  labs(x = "First actor gender", y = "Mean log(budget)")
```

Your Answer:
Write your formulated response here.
step b
```{r}
movies <- movies %>% mutate(first_actor_male = as.numeric(first_actor_gender == "male"))
m3 <- lm(log_budget ~ first_actor_male, data = movies)
summary(m3)
```

Your Answer:
Write your formulated response here.
step c
```{r}
movies %>% filter(first_actor_gender == "female") %>% distinct(first_actor) %>% print(n = Inf)
movies %>% filter(first_actor_gender == "male") %>% distinct(first_actor) %>% print(n = Inf)
```

Your Answer:
Write your formulated response here.
5 Week 5
a. Create a plot of the mean profits by month of release. Do you see any indication that month of release
matters to the profits of the movie? [2 points]
b. Estimate an OLS model which has as dependent variable the log of profits of a movie, and as independent
variable the log of budget, a dummy for whether the movie was released in english or not, and a linear
term for the month of release. Show a summary of the resulting model and interpret each coefficient.
[4 points]
c. Test for the hypothesis that the coefficient that belongs to month of release is zero. [2 points]
d. Based on your plot in a.) do you consider the choice that month of release enters the model linearly
under b.) reasonable? Estimate a specification that allows for a more flexible curve. In this new
specification, test for the null hypothesis that month of release does not impact profits. This might
require testing multiple terms at once. [4 points]
8
e. One executive at the studio wants to time the release of the movie to a specific month of the year such
that they can maximize revenue. Based on your model under d.), What would you advise the movie
studio regarding the timing of the release of the movie? [2 points]
The movie studio that you work at is releasing a new movie in 2026. It will be an English-spoken Thriller
movie with a budget of 40,000,0000.
f. Estimate a model that is able to predict the revenue of this movie. Give its predicted revenue and
include a 99% prediction interval. [6 points]
step a
```{r}
library(tidyverse)
movies <- read_tsv("movies5.tsv") %>%
  mutate(profit = revenue - budget)

movies %>%
  group_by(release_month) %>%
  summarise(mean_profit = mean(profit), se = sd(profit)/sqrt(n())) %>%
  ggplot(aes(factor(release_month), mean_profit)) +
  geom_pointrange(aes(ymin = mean_profit - 1.96*se, ymax = mean_profit + 1.96*se)) +
  labs(x = "Release month", y = "Mean profit")
```

Your Answer:
Write your formulated response here.
step b
```{r}
m <- movies %>% mutate(
  log_profit = ifelse(revenue > 0, log(revenue), NA),
  log_budget = ifelse(budget  > 0, log(budget),  NA),
  english    = as.numeric(original_language == "en"))

m5b <- lm(log_profit ~ log_budget + english + release_month, data = m)
summary(m5b)
```

Your Answer:
Write your formulated response here.
step c
```{r}
summary(m5b)$coefficients["release_month", ]
```

Your Answer:
Write your formulated response here.
step d
```{r}
m5d <- lm(log_profit ~ log_budget + english + factor(release_month), data = m)
m5d_restricted <- lm(log_profit ~ log_budget + english, data = m[!is.na(m$log_profit) & !is.na(m$log_budget), ])
anova(m5d_restricted, m5d)          # joint F-test of all 11 month dummies
```

Your Answer:
Write your formulated response here.
step e
#WRITE YOUR CODE HERE
Your Answer:
Write your formulated response here.
step f
```{r}
m_pred <- m %>% mutate(thriller = as.numeric(genre == "Thriller"))
mf <- lm(log_profit ~ log_budget + english + thriller, data = m_pred)
summary(mf)

new <- data.frame(log_budget = log(40000000), english = 1, thriller = 1)
pr  <- predict(mf, new, interval = "prediction", level = 0.99)
exp(pr)                       # back to dollars
```

